# Dyson's Brownian Motion

We need to simulate an $n \times n$ Hermitian matrix
whose upper triangular entries are independent Brownian Motions,
(with diagonal having twice the variance as the offdiagonal)
and watch how the eigenvalues move.

Here's simulation of *one* matrix:
```{r dbm_init}
efn <- function (B) { eigen(B, symmetric=TRUE, only.values=TRUE)$values }

Binit <- function (n) {
    B0 <- matrix(NA, nrow=n, ncol=n)
    diag(B0) <- rnorm(n) / sqrt(2)
    ut <- upper.tri(B0, diag=FALSE)
    lt <- lower.tri(B0, diag=FALSE)
    B0[ut] <- rnorm(sum(ut))/sqrt(2) + 1i * rnorm(sum(ut))/sqrt(2)
    B0[lt] <- Conj(t(B0)[lt])
    return(B0)
}

B0 <- Binit(10)
stopifnot(all(Mod(B0 - t(Conj(B0))) <= 1e-8))
efn(B0)
```

Here's an update function:
```{r update}
dB <- function (B, dt) {
    B1 <- Binit(nrow(B)) * sqrt(dt)
    return(B + B1)
}

B1 <- dB(B0, 1)
stopifnot(all(Mod(B1 - t(Conj(B1))) <= 1e-8))
efn(B1)
```

Now let's run this for a few steps:
```{r anim_ten}
nsteps <- 100
dt <- 0.1
Blist <- list(Binit(10))
for (k in 2:nsteps) {
    Blist[[k]] <- dB(Blist[[k-1]], dt)
}
Emat <- sapply(Blist, efn)
matplot(t(Emat), type='l', xlab='time', ylab='eigenvalue')
```

Sanity check: the entries of the matrix should be stationary after dividing by $\sqrt{t}$.
```{r check_stationary}
tt <- seq(1, dt*nsteps, length.out=nsteps)
X <- sapply(Blist, "[", upper.tri(Blist[[1]], diag=TRUE))

layout(1:2)
matplot(tt, t(Re(X))/sqrt(tt), type='l', main='Real part')
matplot(tt, t(Im(X))/sqrt(tt), type='l', main='Imaginary part')
```

How big can we do?
```{r speed}
nvals <- 100 * (1:10)
speeds <- lapply(nvals, function (n) system.time(efn(Binit(n))))
plot(nvals, sapply(speeds, "[", 1), type='b',
     xlab='number of rows', ylab='computation time')
```

Let's do with $n=1000$:
```{r anim_more}
nsteps <- 120
dt <- 0.1
Blist <- list(Binit(1000))
for (k in 2:nsteps) {
    Blist[[k]] <- dB(Blist[[k-1]], dt)
}
Emat <- sapply(Blist, efn)

matplot(seq(1, dt*nsteps, length.out=nsteps),
        t(Emat), type='l', xlab='time', ylab='eigenvalue')
```

So, the eigenvalues should scale with $\sqrt{t}$ also.
This should be stationary, but it doesn't look like it.
```{r fluctuations}
tt <- seq(1, dt*nsteps, length.out=nsteps)
E_rescaled <- sweep(Emat, 2, sqrt(tt), "/")

layout((1:2))
matplot(seq(1, dt*nsteps, length.out=nsteps),
        t(E_rescaled)[,1:20], type='l', xlab='time', ylab='eigenvalue',
        main='top 20, sqrt-rescaled')
matplot(seq(1, dt*nsteps, length.out=nsteps),
        t(E_rescaled)[,nrow(Emat)/2+(1:20)-10], type='l', xlab='time', ylab='eigenvalue',
        main='middle 20, sqrt-rescaled')
```


