```{r setup}
set.seed(271)
```

# Brownian motion

First let's start by simulating Brownian motion,
as the scaling limit of simple random walks.

First let's simulate a simple random walk.
```{r sim_srw}
n <- 1002
coins <- sample(c(-1,1), n, replace=TRUE)
plot(cumsum(coins), type='l')
```

Let's look at how this rescales to look like Brownian motion.
```{r bm_sim}
srw_bm <- function (tmax, N) {
    coins <- sample(c(-1,1), floor(tmax * N), replace=TRUE)
    return(cumsum(coins) / sqrt(N))
}
tmax <- 3
Nvals <- c(10, 40, 160, 640)
for (N in Nvals) {
    srw <- srw_bm(tmax, N=N)
    tvals <- seq(0, tmax, length.out=length(srw))
    if (N == Nvals[1]) {
        plot(tvals, srw, type='l', xlab='time', ylab='space', ylim=c(-2*sqrt(3),2*sqrt(3)))
    } else {
        lines(tvals, srw, col=match(N, Nvals))
    }
}
```

Let's test this is working right.
If this is a good approximation to Brownian motion,
the distribution at time 3 should be Normal(0,$\sigma^2=3$).
```{r test_srw}
srw_reps <- replicate(200, srw_bm(3, 10000))
qqnorm(srw_reps[nrow(srw_reps),])
qqline(srw_reps[nrow(srw_reps),])
```
That looks about right. And the sample variance is `r var(srw_reps[nrow(srw_reps),])`,
which is close to 3.


## Function space

We will now simulate Brownian motion on $[0,1]$ using the Levy construction,
using the integrals of the Haar basis and the isometry of $L^2([0,1])$.
```{r haar_basis}
haar_integral <- function (x, n, k) {
    a <- 2 * k / 2^n
    b <- (2 * k  + 1)/ 2^n
    c <- (2 * k  + 2)/ 2^n
    2^((n-1)/2) * ifelse(x < a, 0,
                         ifelse(x < b, x - a,
                                ifelse(x < c, c - x, 0)))
}

curve(haar_integral(x, n=3, k=1), from=0, to=1, ylab='Schauder function value')
curve(haar_integral(x, n=5, k=6), from=0, to=1, add=TRUE, col='red')
curve(haar_integral(x, n=5, k=12), from=0, to=1, add=TRUE, col='blue')
```

Ok, that's our (integrated) basis functions.
Now we need to add them up with independent Gaussian coefficients.
```{r haar_bm}
haar_bm <- function (nmax, bridge=FALSE) {
    if (bridge) {
        nk <- data.frame(n=1, k=0)
    } else {
        nk <- data.frame(n=c(0,1), k=c(0,0))
    }
    for (nn in seq(2, nmax)) {
        nk <- rbind(nk, data.frame(n=nn, k=0:(2^(nn-1))))
    }
    fns <- lapply(1:nrow(nk), function (j)
                  function (x) haar_integral(x, n=nk$n[j], k=nk$k[j]))
    # the random coefficients
    Z <- rnorm(nrow(nk), 0, sd=1)
    return(function (x) {
        Z %*% do.call(rbind, lapply(fns, function (f) f(x)))
    })
}

B <- haar_bm(10)
curve(B, from=0, to=1)
lines(seq(0,1,length.out=1e3), B(seq(0,1,length.out=1e3)), col='red')
```
